Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-12 paper/ms.tex
add Nakajima+1985
ms: improve intro

Data from K2 and TESS have yielded a new class of variable star whose
root cause is only beginning to become clear: complex quasiperiodic
variables (CQVs).  These objects are identified from their optical
light curves, which show nearly periodic troughs that are either sharp
or broad; these troughs are often superposed on smooth spot-like
modulation
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight local minima (``dips'') per cycle.  Most
are pre-main-sequence M dwarfs without near-infrared excesses, ages of
$\approx$5-150 million years (Myr), and rotation periods of at most
two days; they are observed to comprise $\approx$1-3\% of M dwarfs
younger than 100 Myr \citep{2016AJ....152..114R,2022AJ....163..144G}.
The dips can be chromatic, with a reddening law plausibly consistent
with dust
\citep{2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
And finally, while the dip shapes can ``jump'' between different
depths and durations over less than one cycle, they more often evolve
gradually, over tens to hundreds of cycles
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

A few competing explanations for what causes the complex quasiperiodic
variability are shown in Figure~\ref{fig:f1}.  All young M dwarfs are
spotted, which produces flux variations over characteristic timescales
of the rotation period, $P_{\rm rot}$, and its half-harmonic,
$0.5\,P_{\rm rot}$.  The observed dips occur over durations as short
as $0.05\,P_{\rm rot}$; a ``starspot-only'' scenario can be discarded
for any object with sufficiently sharp dips
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The timescales and
amplitudes of the flux variations instead require sharp geometries
with material extrinsic to the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  The
``clump'' scenario invokes opaque dust clumps orbiting near the
Keplerian corotation radius, which periodically eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The ``prominence''
scenario invokes long-lived condensations of cool and dense
marginally-ionized gas, embedded within the hotter corona, that would
be centrifugally supported near corotation
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
Such structures are analogous to quiescent prominences and filaments
seen in the solar corona \citep[see e.g.][]{2015ASSL..415.....V},
though at much larger relative distances from the stellar surface.  A
final possibility that has been suggested is one of a ``screen'', in
which the inner wall of a quiescent circumstellar disk blocks a
portion of the stellar surface to produce sudden dips whenever spots
come into view \citep{2019ApJ...876..127Z}.

@@

		\vspace{-0.6cm}
		\subfloat{
			\includegraphics[width=0.8\textwidth]{f1b.pdf}
		}
	\end{center}
		\vspace{-0.3cm}
	\caption{
		{\bf Complex quasiperiodic variables (CQVs)}:
    {\it Top:} Phase-folded TESS light curves of three CQVs.  Each is
    stacked over one month.  Gray circles are raw 2-minute data; black
    circles bin to 300 points per cycle.  Periods in hours are in the
    bottom right of each panel.  Left-to-right, the objects are LP
    12-502 (TIC 402980664; Sector~19), TIC 94088626 (Sector 10), and
    TIC 425933644 (Sector~28).
    {\it Bottom:} Cartoon models for the phenomenon.  The dust clump
    scenario (lower left) and prominence scenario (lower center) both
    invoke centrifugally-supported material at the corotation radius.
    We disfavor the screen scenario (see Section~\ref{sec:intro}).
	}
	\label{fig:f1}
\end{figure*}

@@

An instructive point of comparison is the rapidly rotating magnetic B
star, $\sigma$~Ori~E, which shows dips that are deeper in the blue
than in the red \citep{1977ApJ...216L..31H}.  Photometric and
spectroscopic observations of this star have been understood in terms
of a warped torus of circumstellar material, analogous to the
geometries we are discussing for much cooler M dwarfs
\citep{1978ApJ...224L...5L,2005ApJ...630L..81T}.  This material
however is unlikely to be dust, due to the relevant sublimation
timescales.  The opacity is instead thought to be sourced from
bound-free absorption from neutral hydrogen
\citep{2005ApJ...630L..81T}, although to our best knowledge direct
evidence for this conclusion has yet to be acquired.  Separate and
smaller-amplitude emission in $\sigma$~Ori~E may also come from
electrons scattering photospheric light toward the observer when the
clouds are not in transit \citep{2022MNRAS.511.4815B}.

@@

%\subsubsection{Why corotation?}
%All of the prominence models, and the diagrams of where the magnetic
%fixed points exist, enable the fixed points to not be located at
%corotation.  In addition, the Balmer absorption line measurements
%have been used to argue for material being generally located at
%positions other than just corotation (CHECK Collier-Cameron1989).

>>>>

Data from K2 and TESS have yielded a new class of variable star whose
root cause is only beginning to become clear: complex quasiperiodic
variables (CQVs).  These objects are identified from their optical
light curves, which show nearly periodic troughs that are either sharp
or broad; these troughs are often superposed on smooth spot-like
modulation
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight local minima (``dips'') per cycle.  Most
are pre-main-sequence M dwarfs without near-infrared excesses, ages of
$\approx$5-150 million years (Myr), and rotation periods of at most
two days; they are observed to comprise $\approx$1-3\% of M dwarfs
younger than 100 Myr \citep{2016AJ....152..114R,2022AJ....163..144G}.
The dips can be chromatic, with a reddening law plausibly consistent
with dust
\citep{2017PASJ...69L...2O,2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
And finally, while the dip shapes can ``jump'' between different
depths and durations over less than one cycle, they more often evolve
gradually, over tens to hundreds of cycles
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

The CQV light curves cannot be explained by pure starspots.  Although
all young M dwarfs are spotted, this tends to produce flux variations
over characteristic timescales of the rotation period, $P_{\rm rot}$,
and its half-harmonic, $0.5\,P_{\rm rot}$.  If one fine-tunes the
viewing geometry, starspots can produce dips with durations as short
as $\approx$$0.2\,P_{\rm rot}$, and the dip amplitudes quickly
decrease due to limb-darkening (see \citealt{2017AJ....153..152S},
Figures 37-41).  The observed dips in CQVs can happen over durations as
short as $0.05\,P_{\rm rot}$ with characteristic depths of a few
percent; a ``starspot-only'' scenario can be discarded for any object
with dips that are sufficiently deep and short in duration
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The timescales and
amplitudes of the flux variations in such objects instead require
sharp geometries with material extrinsic to the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  Since the
entire class of star is probably part of the same phenomenon, we
therefore generally discard the ``spot-only'' scenario.

A few competing explanations for what causes the complex quasiperiodic
variability are shown in Figure~\ref{fig:f1}.  The ``clump'' scenario
invokes opaque dust clumps orbiting near the Keplerian corotation
radius, which periodically eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The ``prominence''
scenario invokes long-lived condensations of cool and dense
marginally-ionized gas, embedded within the hotter corona, that would
be centrifugally supported near corotation
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
Such structures are analogous to quiescent prominences and filaments
seen in the solar corona \citep[see e.g.][]{2015ASSL..415.....V},
though at much larger relative distances from the stellar surface.  A
final possibility that has been suggested is one of a ``screen'', in
which the inner wall of a quiescent circumstellar disk blocks a
portion of the stellar surface to produce sudden dips whenever spots
come into view \citep{2019ApJ...876..127Z}.

@@

		\vspace{-0.6cm}
		\subfloat{
			\includegraphics[width=0.8\textwidth]{f1b.pdf}
		}
	\end{center}
		\vspace{-0.3cm}
	\caption{
		{\bf Complex quasiperiodic variables (CQVs)}:
    {\it Top:} Phase-folded TESS light curves of three CQVs.  Each is
    stacked over one month.  Gray circles are raw 2-minute data; black
    circles bin to 300 points per cycle.  Periods in hours are in the
    bottom right of each panel.  Left-to-right, the objects are LP
    12-502 (TIC 402980664; Sector~19), TIC 94088626 (Sector 10), and
    TIC 425933644 (Sector~28).
    {\it Bottom:} Cartoon explanations for the phenomenon.  The dust clump
    scenario (lower left) and prominence scenario (lower center) both
    invoke centrifugally-supported material at the corotation radius.
    We disfavor the screen scenario (see Section~\ref{sec:intro}).
	}
	\label{fig:f1}
\end{figure*}

@@

An instructive point of comparison is the rapidly rotating magnetic B
star, $\sigma$~Ori~E, which shows dips that are deeper in the blue
than in the red \citep{1977ApJ...216L..31H}.  Photometric and
spectroscopic observations of this star have been understood in terms
of a warped torus of circumstellar material, analogous to the
geometries we are discussing for much cooler M dwarfs
\citep{1978ApJ...224L...5L,1985Ap&SS.116..285N,2005ApJ...630L..81T}.  This material
however is unlikely to be dust, due to the relevant sublimation
timescales.  The opacity is instead thought to be sourced from
bound-free absorption from neutral hydrogen
\citep{1985Ap&SS.116..285N}, although to our best knowledge direct
evidence for this conclusion has yet to be acquired.  Separate and
smaller-amplitude emission in $\sigma$~Ori~E may also come from
electrons scattering photospheric light toward the observer when the
clouds are not in transit \citep{2022MNRAS.511.4815B}.

@@

%\subsubsection{Why corotation?}
% The stable point only exists if R_A > R_K.  For radii between R_A and R_K in such cases, consider a magnetic field line extending out beyond R_K.  Whenever R>R_K, the material is centrifugally supported (pushed out by the ~Ω^2R^2 centrifugal term in the corotating frame).  If R<R_K in such cases along a given field line, the material will fall back to the star.  So there are two conditions that must be met: the star must spin fast enough, and closed magnetic loops must extend beyond the corotation radius.  (This is all spelled out by Nakajima1985, and Petit2013).
